---
title: "Suicide Rates"
output:
  github_document:
    toc: true
    toc_depth: 3
editor_options:
  chunk_output_type: inline
---
# Suicide Rates of Men and Women

## Libraries

```{r}
suppressMessages(library(tidyverse))
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(rworldmap))
suppressMessages(library(corrplot))
suppressMessages(library(modelr))
```


## The Data
The data I will be using are from the World Health Organization:

**THE DATA:**
Some data has been removed due to lacking a substantial amount of inputs or causing problems with how it was originally set up(HDI, data from 2016, and the generation variable)

**country:** Country the suicide is in

**year:** year it occurred

**sex:** gender of the individual

**age:** age of the individual
  split up into groups
  5-14
  15-24
  25-34
  35-54
  55-74
  75+

**suicides_no:** number of suicides in that year for that sex and age group 

**population:** population of the country for that year, sex, and age

**suicides.100k.pop:** suicides per 100k population per gender per age group

**gdp for year:** Gross Domestic Product for the year

**gdp per capita:** Gross Domestic Product per capita

## The Study

The main purpose of analyzing global suicides from 1958 to 2016 is to answer a few questions that came to mind when seeing this data; How has the rate of suicide has changed over time, which gender is more likely to commit suicide and the level of impact each factor has on it, is there a correlation between GDP and the suicide rates of countries, and do different countries have different trends in suicides. Another thought I had when finding this data, was to see if there are trends and when similar data is released from 2020, during the Covid pandemic it will be possible to compare the past trends with it and see how the patterns have changed.

**The main questions I'll be answering and expounding upon are as follows:**

**1.** How has the rate of suicide changed over time?

**2.** Which gender is more likely to commit suicide?

**3.** Is there a correlation between GDP and the suicide rates of countries?
        - Do different countries have different rates of suicide?


```{r}
#loading in data
df<- read.csv("master.csv")
head(df)
colnames(df)

```
## Data Cleaning


```{r}
#cleaning up the data -- removing hdi and generation and country.year as a lot of entries had no value for HDI and the generation value causes problems with the continuous data
df <- df[ !(grepl("2016", df$year) ), ]
drops <- c("HDI.for.year", "generation", "country.year")
df <-df[ , !(names(df) %in% drops)]

df <- df %>%
  rename(gdp_for_year = gdp_for_year...., 
         gdp_per_capita = gdp_per_capita....)

#create new column to have age as a numeric value
df <- df %>%
        mutate(ages = as.numeric(factor(age)))
head(df)
```


## Analysis


### The Change in Suicide Rate Over Time
          
If we first look at the amount of suicides that have occurred each year from 1985-2015 we can get a general idea about the change over time. Then, go more into specific variables and their changes over time in regards to suicides per 100k later on to answer the question about how the rate of suicide has changed over time and some other related variables along with it.


```{r}
#show the trend of global suicides per 100k over time

df %>% 
  group_by(year) %>%
  summarise(Suicides = mean(sum(suicides_no))) %>%
  ggplot(aes(year, Suicides)) + 
  geom_line(col= "red") + 
  geom_point(col = "blue") +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  scale_y_continuous(breaks = seq(50000, 300000, 25000), labels = scales::comma) +
  theme(plot.caption=element_text(hjust = 0)) +
  
  labs(title = "Global Suicides", 
       subtitle = "Total suicides from 1985-2015",
       x = "Gender",
       y = "Suicides",
       caption = "• The amount of suicides around the world was increasing up until around 2002 and has since decreased 
       even with an ever-growing global population size.
• The most suicides occurred in 2002 with 256,095.
• In 2015 there were 203,640 suicides with a recorded population of roughly 1.8 billion. ")


df %>% 
  group_by(year) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop))) %>%
  ggplot(aes(year, suicides100k)) + 
  geom_line(col= "red") + 
  geom_point(col = "blue") +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  scale_y_continuous(breaks = seq(10, 15, 1)) +
  theme(plot.caption=element_text(hjust = 0)) +
  
  labs(title = "Global Suicides", 
       subtitle = "suicide rate per 100k from 1985-2015",
       x = "Year",
       y = "Rate of suicides per 100k",
       caption = "• The rate of suicides per 100k was steadily increasing with a big spike from 1994-1995 and has been decreasing 
       since. 
• The highest rate of suicides (per 100k) was in 1995 with roughly 15.7. 
• In 2015 the rate of suicides (per 100k) was about 11.1.")

<<<<<<< HEAD

=======
A pretty steep increase in suicides per 100k up until the big spike from 1994-1995, then overall decreasing with some small increases between years
>>>>>>> 637dbfde8b11e0645d1e9763fa644e110db65243

df %>% 
  group_by(year) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop)))

df %>%
  group_by(year) %>%
  summarize(sum(population))

df %>% 
  group_by(year) %>%
  summarise(Suicides = mean(sum(suicides_no)))

```
A pretty steep increase in suicides per 100k up until the big spike from 1994-1995, then overall decreasing with some small increases between years


We can also look at the change over time with the rate of suicides to get an idea of how rapidly it’s decreasing or increasing. This will also give a base answer to my question of how the rate of suicides (per 100k) has changed over time, with a few more in-depth graphs with more variables included to follow it.

As a whole, the rate of suicides (per 100k) has been getting better (lower/decreasing), of course the rate is based on the data available and could differ slightly or majorly depending on how similar it is to the populations not included. I wouldn’t expect it to differ majorly if every single suicide was able to be recorded but of course there’s always the possibility that it could.



### Differences in Gender with Suicide

Next, is the question of whether one gender is more likely to commit suicide and if this varies globally. I’ll analyze this at the basic level and then include a few different styles of graphs with different variables included to help display the difference in suicide rates (per 100k) and total suicides between the two genders.

First, let’s look at the overall rate of suicides (per 100k) divided by sex.

I’ve also created a line graph with the suicide rates (per 100k) separated by gender to see how it’s changed over time.


```{r}
#bar graph showing rate of suicides per 100k by sex
df %>% 
  group_by(sex) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop))) %>%
  ggplot(aes(sex, suicides100k, fill=sex)) + 
  geom_bar(stat="identity", col= "black") +
  scale_y_continuous(breaks = seq(0, 22, 2)) +
  theme(plot.caption=element_text(hjust = 0)) +

labs(title = "Global Suicides by Gender", 
       subtitle = "suicide rate per 100k from 1985-2015",
       x = "Gender",
       y = "Rate of suicides per 100k",
       caption = "• Males’ suicide rate (per 100k) is roughly 4 times greater than that of females.
• Males’ suicide rate (per 100k) is about 20 while females is 5.4.")

#showing the trend of each sex's suicides over
df %>% 
  group_by(year, sex) %>%
  summarise(suicides100k = mean(suicides.100k.pop)) %>%
  ggplot(aes(x=year, y=suicides100k, col=factor(sex))) + 
  geom_line() + 
  geom_point()+
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  scale_y_continuous(breaks = seq(0, 24, 2)) +
  theme(plot.caption=element_text(hjust = 0)) +

  labs(title = "Global Suicides by Gender", 
       subtitle = "suicide rate per 100k from 1985-2015",
       x = "Year",
       y = "Rate of suicides per 100k",
       caption = "• Males’ suicide rate (per 100k) has a much higher base rate and has increased up until 1995 and has 
       steadily been decreasing to a similar rate of 1987.
• Females’ suicide rate (per 100k) has been steadily decreasing since 1985 and hasn’t had any spikes")

#getting numbers of these graphs for comparison purposes
df %>% 
  group_by(country) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop)))

df %>%
  group_by(sex) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop)))
  
```


With the overall rate of suicides per 100k vs. year, (Figure 2) the large
spike from 1994-1995 seems to have mostly been caused my the male rate of
suicides (per 100k) increasing at a considerably greater rate than average.



### Geographical Analysis

Now to check and see if the rate of suicides (per 100k) varies around the world,
and if the rate for each gender is also similar. To do this, we will look at a heatmap to see the rate of suicides (per 100k) around the world for the countries in the data set.


```{r}
#showing global heatmap of suicides
continent <- df %>%
  group_by(country) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop)))


mapped_data <- joinCountryData2Map(continent, joinCode = "NAME", 
                                   nameJoinColumn = "country")

par(mar=c(0,0,0.5,0), xaxs="i", yaxs="i")

#graphing
mapCountryData(mapped_data, nameColumnToPlot = "suicides100k",
               missingCountryCol = "white",
               oceanCol = "lightblue", mapTitle = "Global suicides per 100k"
               )
```

There is a much higher rate of suicides (per 100k) in the European and Asian regions.


This heat-map shows that the rate of suicides (per 100k) in European and Asian regions is about double that of North America and Australia. It’s hard to confirm that this is the exact spread of suicide rates (per 100k) globally as there’s practically no data for Africa, little for South America, and even less for Southern Asia. But, with the data at hand this does allow us to see the rate of suicides (per 100k) visually around the globe.


We then get the rate of suicides (per 100k) for each country and visually break it up between male and female to see if there’s a difference in the rate of suicides (per 100k) for each gender around the globe.


### Country Analysis

     
```{r, fig.height= 13}
#showing country breakdown by sex
countries <- df %>% 
  group_by(country, sex) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop)))

#graph
  ggplot(countries, aes(x = reorder(x = country, suicides100k), suicides100k, fill = sex)) + 
  geom_bar(stat="identity") +
  coord_flip() +
  theme(plot.caption=element_text(hjust = 0)) +

  labs(title = "Suicides by Country", 
       subtitle = "suicide rate per 100k from 1985-2015",
       x = "Country",
       y = "Rate of suicides per 100k", 
       caption = "• The ratios between the genders seem to be fairly similar between countries 
       throughout the world with a couple having slightly higher female suicide 
        rates (per 100k) than average and some having slightly lower.
• Lithuania appears to have the highest rate of suicides (per 100k), with 68.14 for males, 
roughly 5 times the rate (per 100k) of females, with 13.03.")
  
  #to get the numbers of it
df %>%
  group_by(country, sex) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop)))
```


This is also not a 100% accurate conclusion, as without data from Africa or the countries missing, we’re uncertain if there could be differences in the rates (per 100k) of each gender of those countries, and different areas around the world could have different classifications of suicides as well.


## Model 

The main question I want to answer with this research is whether a country’s annual GDP has any effect on its suicide rate. If so, in what way? Will poorer countries have a higher rate of suicides or richer? The one downside to this data set is it’s lack of data for African countries, typically on the lower end of the global GDP scale, so the results of this model could be incorrect when applied with data from every country in the world. I start with grouping up the data by country, then getting the average of the suicides per 100k and adding these all together to get the countries suicide rate per 100k from 1985-2015 into one value for each country. Then, getting the GDP per capita and GDP for the year similarly.


```{r}
#setting up df for regressions
byCountry <- df %>%
  group_by(country) %>%
  summarise(suicides100k = sum(mean(suicides.100k.pop)), 
            gdpCapita = mean(gdp_per_capita),
            gdpYear = mean(gdp_for_year
            ))

#plotting residuals
ggplot(byCountry, aes(x = gdpCapita, y = suicides100k)) + 
  geom_point()

ggplot(byCountry, aes(x = gdpYear, y = suicides100k)) + 
  geom_point() + scale_x_continuous(limits = c(0, 150000000000), labels = scales::comma)

  ggplot(byCountry, aes(x = gdpCapita, y = suicides100k)) + 
  geom_point() +
  geom_smooth() + xlim(0,40000) +ylim(0,30) +
  theme(plot.caption=element_text(hjust = 0)) +

  labs(title = "Correlation between Suicides per 100k and GDP per Capita", 
       x = "GDP per Capita",
       y = "Rate of suicides per 100k", 
       caption = "• Again, with the rate of suicides (per 100k) plotted against the GDP per capita, there isn’t much of a pattern or 
       correlation between the two variables.")

model1 <- lm(suicides100k ~ gdpCapita + gdpYear, byCountry)
summary(model1)

model2 <- lm(suicides100k ~ gdpYear, byCountry)
summary(model2)


ggplot(byCountry, aes(x = gdpYear, y = suicides100k)) + 
  geom_point() +
  geom_smooth() +
  scale_x_continuous(limits = c(0, 150000000000), labels = scales::comma) +
  theme(plot.caption=element_text(hjust = 0)) +
  
labs(title = "Correlation between Suicides per 100k and GDP", 
       x = "GDP",
       y = "Rate of suicides per 100k", 
       caption = "• As visible in the graph, the best fit line can’t explain much of the varying rates of suicide (per 100k) with different 
     levels of GDP.")




#checking if mean of residuals is centered around 0

canned.ols <- lm(suicides.100k.pop ~ gdp_per_capita + gdp_for_year, df)
summary(canned.ols)

datadata <- df %>% 
  add_predictions(canned.ols) %>% 
  add_residuals(canned.ols)
head(datadata)


datadata %>% 
  summarize(mean_of_resid = mean(resid))

```


Since the P values, Pr(>|t|), are both greater than .05, .846 and .460, respectively, we fail to reject the null hypothesis that a country’s GDP per capita and annual GDP have no association with it’s rate of suicides (per 100k). The r-squared is .007374 so annual GDP and GDP per capita don’t explain much of the variance in the suicide rates (per 100k). We can see this visualized on the graphs above.






## Conclusions and Final Thoughts


The original assumptions I had for the questions that I proposed didn’t exactly match up with the answers I concluded from the data. The suicide rate (per 100k) has decreased over time at a pretty substantial rate with males being the ultimate driving force of this decrease as they are roughly 4 times more likely to commit suicide than females. The suicide rates (per 100k) have similar distributions between the two sexes around the world. With the model used, no correlation between the GDP a country has in a given year, or its GDP per capita, was found with its rate of suicides.
